Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-20 doc/network.tex
minor edits by DC

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
\jd{dA is better, and we obviously need to clarify.}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.

@@

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

Generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease.

@@

We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
\end{equation}
where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution. 
This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.
\swp{This is using epigrowthfit. How do I cite it? It's not out yet...}
\dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?}

>>>>

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
\jd{dA is better, and we obviously need to clarify.}
\dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.

@@

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.} 
\dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

Generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.

@@

We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
\end{equation}
where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution. 
This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.
\swp{This is using epigrowthfit. How do I cite it? It's not out yet...}
\dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?} \dc{Oh OK, yes}
